## DATA WRANGLING:        
- process of reimaging the raw data to a more structured format, which will help to get better insights and make better decisions from the data.
- dplyr() package for data wrangling that will allow us to take a data frame and “wrangle” it (transform it) to suit our needs.


#### Recursive object & function ??????????  :
-recursive object in R is an object that can contain another object of the same type as itself.         
-It's like a structure that repeats or nests within itself.       

-Imagine you have a box, and inside that box, there's another identical box.        
-Now, if you open the second box, you might find a third identical box, and this pattern can continue.       

-In R, this concept is often seen in data structures like lists or environments.        
-For example, you can have a list where one of its elements is another list, and within that sublist, there might be another list, and so on.



*Subsetting:*       
-subsetting allows the user to access elements from an object.        
-takes out a portion from the object based on the condition provided.       

* Method 1: Using [] operator:        
```{r}
#creating a vector:
x <- 1:20
# subsetting a vector:
cat("First 5 values of vector:",x[1:5],"\n")
cat("WIthout values present at index 1,2,3:",x[-c(1,2,3)])
```

```{r}
#dataset:
cat("dataset:" ,"/n")
print(head(mtcars))
#subsetting:
cat("hp values of all cars:", "/n")
print(head(mtcars["hp"]))

```

* Method 2: Using [[]] operator:        
-used for subsetting of list-objects.       
-difference is that [[ ]] selects only *one element* whereas [ ] operator can select *more than 1 element* in a single command.        

```{r}
# Create list
ls <- list(a = 1, b = 2, c = 10, d = 20)

# Print list
cat("Original List: \n")
print(ls)

# Select first element of list
cat("First element of list: ", ls[[1]], "\n")

```

```{r}
#Eg2
# Create list
z <- list(a = list(x = 1, y = "GFG"), b = 1:10)

# Print list
cat("Original list:\n")
print(z)

# Print GFG using c() function
cat("Using c() function:\n")
print(z[[c(1, 2)]])

# Print GFG using only [[]] operator
cat("Using [[]] operator:\n")
print(z[[1]][[2]])

```

* Method 3: Using $ operator:       
-used for lists and data frames in R.       
-Unlike [ ] operator, it selects only a single observation at a time.       
-can be used to access an element in named list or a column in data frame.        
-$ operator is only applicable for recursive objects or list-like objects.        
    
```{r}
# Create list
ls <- list(a = 1, b = 2, c = "Hello", d = "GFG")

# Print list
cat("Original list:\n")
print(ls)

# Print "GFG" using $ operator
cat("Using $ operator:\n")
print(ls$d)
```
```{r}
# Dataset
cat("Original data frame:\n")
print(mtcars)

# Access hp column
cat("Using $ operator:\n")
print(mtcars$hp)

```


* Method 4: Using subset() Function :               
-subset() function in R programming is used to create a subset of vectors, matrices, or data frames based on the conditions provided in the parameters.        
-Syntax: subset(x, subset, select)        
Parameters: 
x: indicates the object
subset: indicates the logical expression on the basis of which subsetting has to be done
select: indicates columns to select

```{r}
# Subsetting
airq <- subset(airquality, Temp < 65, 
				select = c(Month))

# Print subset
print(airq)

```
```{r}
# Subsetting
mtc <- subset(mtcars, gear == 5 & hp > 200, 
					select = c(gear, hp))

# Print subset
print(mtc)

```
















*Tibbles* :
- Core data structure of the tidyverse and are used to facilitate the display and analysis of information in a tidy format. 
-similar to a data frame( or new form of df: where data frames are the most common data structures used to store data sets in R) but comes with some enhanced features and improvements. 
```{r}
# Load the tidyverse package
library(tidyverse)

# Create a tibble
my_tibble <- tibble(
  Name = c("Alice", "Bob", "Charlie"),
  Age = c(25, 30, 22),
  Score = c(95, 89, 75)
)

# Print the tibble
print(my_tibble)

df <- data.frame(
  fruits = c("apple" ,"oranges" ,"grapes"),
  color = c("red","orange","green")
)
print(df)
```
**Advantages:**
* All Tidyverse packages support Tibbles.
* Tibbles print in a much cleaner format than data frames.
* A data frame often converts character strings to factors and analysts often have to override the setting while Tibbles doesn’t try to make this conversion automatically.


* 1.as_tibble():         
-The first function is as tibble function.        
-This function is used to create a tibble from an *existing data frame*.        
-Syntax: as_tibble(x, validate = NULL, …) x is either a data frame, matrix, or list.        

* 2.tibble():        
-The second way is to use a tibble() function, which is used to create a *tibble from scratch.*       
-Syntax: tibble(s…, rows = NULL) s represents a set of name-value pairs.       

* Import(): Finally, you can use the tidyverse’s data import packages to create Tibbles from external data sources such as databases or CSV files.
    
* library(): The library() function is used to load the namespace of the package.       

```{r}
library(tidyverse)
head(CO2) #data frame
```

```{r}
#Method 1 : Converting to tibble using as_tibble:
sample_tibble <- as_tibble(CO2)
print(sample_tibble)

#Method 2: create a tibble from scratch: using tibble():
name <- c("Pk","Sb","Hg","Sg")
marks_maths <- c(92,91,90,89)
marks_jj <- c(27,29,23,24)
color <- c("red","pink","blue","green")
students <- tibble(name,marks_maths,marks_jj,color)
print(students)
```


* Subsetting tibbles:*       
-Data analysts often extract a single variable from a tibble for further use in their analysis : called as *subsetting*.       
-When we try to subset a tibble, we extract a single variable from the Tibble in vector form.       
-Use $ and [[]] operator .       

* Extracting(access a single variable) Using $ operator:*
```{r}
students$color
```

* Extracting using [[ ]]*
```{r}
students[["name"]]
```

* Filtering Tibbles:*       
-Helps to reduce the number of rows in your tibble.
-When filtering, we can specify conditions or specific criteria that are used to reduce the number of rows in the dataset.
-Syntax: filter(data, conditions)

```{r}
filter_students <- filter(students,marks_jj>24)
filter_students
```

* Converting a tibble:*
-Converting a traditional *data frame* to a *tibble*: as_tible
```{r}
data_students <- data.frame(name,marks_maths,marks_jj,color)
data_students
tibble_data_converted <- as_tibble(data_students)
tibble_data_converted
```







































